```{r}
# libraries
library(tidyverse)
library(scales)
```

```{r}
rail_data <- read_csv("./Data/Weekly_Cargo_Data_2020_2026_noUP.csv")
```
Make data columns numeric
```{r}
# transfomr data to numerical
rail_data$Originated <- as.numeric(rail_data$Originated)
rail_data$Received<- as.numeric(rail_data$Received)
```

Drop Empty containers and trailers
```{r}
# drop empty containers and trailers from data
rail_data <- rail_data %>%
  filter(!(`Commodity Group Code` %in% c("U", "V")))
```


Create Normalized Commodity numbers
```{r}
# We group by the columns that define your "n" rows (e.g., Railroad and Week)
rail_data <- rail_data %>%
  group_by(Company, Week) %>% 
  mutate(
    # This creates a column with the total carloads for that specific RR/Week
    Weekly_Total = sum(Originated, na.rm = TRUE),
    
    # This calculates the proportion for every row relative to that total
    Originated_Norm = Originated / Weekly_Total
  ) %>%
  ungroup() # Always ungroup after you're done to avoid logic errors later

# We group by the columns that define your "n" rows (e.g., Railroad and Week)
rail_data <- rail_data %>%
  group_by(Company, Week) %>% 
  mutate(
    # This creates a column with the total carloads for that specific RR/Week
    Weekly_Total = sum(Received, na.rm = TRUE),
    
    # This calculates the proportion for every row relative to that total
    Received_Norm = Received / Weekly_Total
  ) %>%
  ungroup()
```


** KEEP EVAL = FALSE IF YOU DONT WANT 100+ RAW COMMODITY OVER TIME GRAPHS **
lowkey still helpful though

Initial weekly graphs of each Cmmodity Code for each Company
```{r, eval = FALSE}
# 1. Get every unique combination of Railroad and Commodity
report_list <- rail_data %>% 
  distinct(Company, `Commodity Group Code`)

# 2. Loop through the list to print individual plots
for(i in 1:nrow(report_list)) {
  
  # Identify current Railroad and Commodity
  current_rr <- report_list$Company[i]
  current_comm <- report_list$`Commodity Group Code`[i]
  
  # Filter data for just this specific pairing
  plot_data <- rail_data %>%
    filter(Company == current_rr, `Commodity Group Code` == current_comm)
  
  # Create the individual plot
  p <- ggplot(plot_data, aes(x = Week, y = Originated_Norm)) +
    geom_line(color = "steelblue", linewidth = 1) +
    geom_area(fill = "steelblue", alpha = 0.2) +
    theme_minimal(base_size = 12) + 
    labs(
      title = paste("Carrier:", current_rr),
      subtitle = paste("Commodity:", current_comm),
      x = "Date",
      y = "Weekly Originated Volume"
    ) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )

  # 3. Print the plot to the active graphics device (Plots pane)
  print(p)
}
```



```{r}
# get industry wide weekly averages

industry_averages <- rail_data %>%
  # 1. Get the total industry volume for EACH week (to use as denominator)
  group_by(Week) %>%
  mutate(Total_Industry_Cars_This_Week = sum(Originated, na.rm = TRUE)) %>%
  
  # 2. Get the total for EACH commodity across all railroads per week
  group_by(`Commodity Group Code`, Week) %>%
  summarise(
    Weekly_Comm_Sum = sum(Originated, na.rm = TRUE),
    # This is the TRUE proportion of the industry that week
    Industry_Prop_This_Week = Weekly_Comm_Sum / max(Total_Industry_Cars_This_Week),
    .groups = "drop"
  ) %>%
  
  # 3. Average those weekly industry proportions over the whole time period
  group_by(`Commodity Group Code`) %>%
  summarise(
    Final_Avg_Proportion = mean(Industry_Prop_This_Week, na.rm = TRUE)
  )
```

# Get industry wide weekly average proportions for each commodity 
```{r}
# 1. Extract Industry-Wide Proportion for every week
industry_weekly <- rail_data %>%
  group_by(Week) %>%
  mutate(Total_Industry_Vol = sum(Originated, na.rm = TRUE)) %>%
  group_by(`Commodity Group Code`, Week) %>%
  summarise(
    Industry_Prop = sum(Originated, na.rm = TRUE) / max(Total_Industry_Vol),
    .groups = "drop"
  )
```


# Join averages to dataframe
```{r}
# 2. Join this benchmark back to your main data
rail_data <- rail_data %>%
  # Calculate the individual company's proportion first
  group_by(Company, Week) %>%
  mutate(Company_Prop = Originated / sum(Originated, na.rm = TRUE)) %>%
  ungroup() %>%
  # Merge the industry benchmark in
  left_join(industry_weekly, by = c("Commodity Group Code", "Week")) %>%
  # Calculate the Difference (Delta)
  mutate(Prop_Diff = Company_Prop - Industry_Prop)
```


```{r}
# 3. Create the Visualization Loop
# We will iterate through each Railroad/Commodity to see the "Delta"
report_list <- rail_data %>% distinct(Company, `Commodity Group Code`)

# define set lower and upper limits
y_limit_lower <- -0.5
y_limit_upper <- 0.5

for(i in 1:nrow(report_list)) {
  
  curr_rr <- report_list$Company[i]
  curr_comm <- report_list$`Commodity Group Code`[i]
  
  plot_data <- rail_data %>%
    filter(Company == curr_rr, `Commodity Group Code` == curr_comm)
  
  p <- ggplot(plot_data, aes(x = Week, y = Prop_Diff)) +
    # The "Zero" line represents the Industry Average
    geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 1.2) +
    
    # The actual difference over time
    geom_line(color = "midnightblue", linewidth = 1) +
    
    # Optional: Gradient fill to show 'Over' vs 'Under' performance
    geom_area(aes(fill = Prop_Diff > 0), alpha = 0.2, show.legend = FALSE) +
    scale_fill_manual(values = c("TRUE" = "green", "FALSE" = "red")) +
    
    theme_minimal(base_size = 12) +
    
    # CRITICAL: This fixes the scale so all graphs look the same
    coord_cartesian(ylim = c(y_limit_lower, y_limit_upper)) +
    
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    
    labs(
      title = paste("Carrier Relative Performance:", curr_rr),
      subtitle = paste("Commodity Group:", curr_comm),
      caption = "Red Dashed Line = Industry Benchmark (0% Difference)",
      x = "Date",
      y = "Market Share Deviation"
    ) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
  
  print(p)
}
```


# Filter out the most deviant Commodity-Company Pairs
```{r}
# 1. Calculate the "Distance from Zero" Score
dist_from_zero_rank <- rail_data %>%
  group_by(Company, `Commodity Group Code`) %>%
  summarise(
    # Mean Absolute Deviation from the 0 baseline
    avg_distance = mean(abs(Prop_Diff), na.rm = TRUE),
    # Max single-point spike away from 0
    max_deviation = max(abs(Prop_Diff), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Take the Top 10 pairs that are furthest from the industry average
  slice_max(avg_distance, n = 8)

# 2. Filter the main data for these "Distance Outliers"
top_dist_pairs <- rail_data %>%
  mutate(pair_id = paste(Company, `Commodity Group Code`)) %>%
  filter(pair_id %in% paste(dist_from_zero_rank$Company, dist_from_zero_rank$`Commodity Group Code`))

top_dist_pairs$pair_label <- paste(top_dist_pairs$Company,
                                   top_dist_pairs$`Commodity Group`,
                                   sep = ": ")
```

# Plot the top 10 largest differentials
```{r}
# 1. Create a custom color palette where each Railroad has a "family" of shades
# This ensures BNSF lines are all various blues, UP are all various greens, etc.
unique_pairs <- unique(top_dist_pairs$pair_label)
base_colors <- c("blue", "darkorange", "darkgreen", "magenta", "firebrick")
manual_colors <- setNames(
    colorRampPalette(base_colors)(length(unique_pairs)),
    unique_pairs
  )

# 2. Plotting with manual color override
ggplot(top_dist_pairs, aes(x = Week, y = Prop_Diff, color = pair_label)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 1.5) +
  # Opacity set to 0.6
  geom_line(linewidth = 1, alpha = 0.6) +
  
  # Use the manual scale to get those distinct shades per pair
  scale_color_manual(values = manual_colors) +

    # Styling for clarity
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 10) +
  labs(
    title = "Top 8 Commodity-Company Pairs Furthest from Industry Average",
    x = "Date",
    y = "Market Share Deviation",
    color = "Carrier : Commodity"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 8, face = "bold"),
    plot.title = element_text(face = "bold")
  ) 
# TODO: Volitality analysis//clean to combine commodities
```

# save image
```{r}
# Save the final comparison graph
ggsave(
  filename = "Top_10_Market_Deviations.png", 
  plot = last_plot(),       # Captures the very last plot you generated
  width = 12,               # Width in inches
  height = 7,               # Height in inches
  dpi = 300,                # High resolution for reports/presentations
  bg = "white",              # Ensures the background isn't transparent
  device = png
)
```

